Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-03-31 paper/metamonads.tex
section about monads is half-finished.

@@

We will be using MetaOcaml, which, as the instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them

@@

(in the floating-point instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code expressions (perhaps say one more sentence about the brackets).
The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes, which are
...

We can see the result in
\begin{code}
let simplest_code =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

MetaOCaml helpfully prints out the result of the latter, as
|.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that 'gen x y' looks like a combinator; the body of
'gen x y' involves several apoplications of the function 'plus'. When
we look at the generated code however, we see no traces of function
'plus' and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the above code: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
|simplest_code| is not the function on integers -- it's a code
expression.

@@

and use it to generate code that operates on integers, floating pint
numbers or boolean -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a bollean function.

We can see the separation of concerns, implementation of domain
operations.

@@

We see that expression `(plus y one)' appears twice. That may be quite
a complex computation (depending on the implementation of `plus'), and
so we would rather not do it twice. We may be tempted to rely on the
compiler's built-in common-subexpression elimination algorithm, which
is a typical part of the optimizer. We would rather not relied on the
sufficiently smart compiler: when the generated code is very complex,
the compiler may overlook common subexpressions (or the subexpression
occur in such an imperative context that the compiler might not be
able to decide if lifting the subexpression is sound. So, being
conservative, the optimizer will leave the duplicates). We may attempt
to do
\begin{code}
let param_code1 plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

However, examinining the code for |param_code1 plus one|
shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
we might have saved the computation on the code generation
stage. However, the generated code still exhibits the
duplication (which makes the code larger and may take longer to
execute, if the optimizer fails to eliminate the duplication).
So, we need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| so we could use
that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
But that seems impossible because |letgen exp| needs to generate
the let expression .<let t = exp in body>. but |letgen| doesn't have
the body yet. The body needs that temporary identifier .<t>. that will
be generated by |letgen| itself. Certainly |letgen| cannot generate
only a part of a let-expression. All generated expressions in
MetaOCaml are well-formed and complete.

The key is to use CPS. It was first pointed out by XXX in the context
of partial evaluation, and extensively used by (FFT) for code
generation.

@@

Now, |param_code2 plus one| gives us the desired code
|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.

@@

What is our monad: what is the action and what is the value.
ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
the connection with the delimited continuations: for future work.

@@

>>>>

\oleg{explictly enumerate our contributions, not foiregtting the mdo
  notation. The subsubsection that outlines what the other sections
  are for.}

@@

We will be using MetaOcaml, which, as the instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them.

@@

(in the floating-point instance of that aspect, we would have
|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
code expressions -- in other words, to construct future-stage
computations. The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes |.~|. They
let us perform an immediate code generating computation while we are
building the future-stage computation (code expression). In the
example above, the immediate computation |.~x| is trivial -- obtaining
the value of the variable |x|, which should be a code expression, and
inlining it. The following shows an example of using the simple
generators:

\begin{code}
let simplest_code =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
it includes the evaluation of the function |gen|, which in turn
applies the function |plus|. The function receives code expressions
|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
manipulate code expressions as values. The function |gen| returns a
code expression, which will be inlined in the place of the
escape. MetaOCaml can print out code expressions, so we can see the
final generated code:\\
  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that |gen x y| looks like a combinator; its body 
involves several applications of the function |plus|. When
we look at the generated code however, we see no traces of function
|plus| and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the code expression: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
|simplest_code| is not the function on integers -- it is a code
expression.

@@

and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a bollean function.

We can see the separation of concerns, namely of that for domain
operations.

@@

We notice two occurrences of |plus y one|. Depending on the
implementation of |plus| that may be quite a complex computation, and
so we would rather not do it twice. We may be tempted to rely on the
compiler's common-subexpression elimination optimization. We would
rather not relied on a ``sufficiently smart compiler'': when the
generated code is very complex, the compiler may overlook common
subexpressions. The subexpressions may occur in such an imperative
context that the compiler might not be able to determine if lifting
them is sound. So, being conservative, the optimizer will leave the
duplicates as they are. We may attempt to write

\begin{code}
let param_code1 plus one =
  let gen x y = 
     let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

However, the result of |param_code1 plus one|, which prints as\\
|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
still exhibits duplicate sub-expressions. Our |let|-insertion
optimization saved the computation on the generating stage. 
We need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| to be used
as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
yielding the code like |.<let t = y + 1 in t + (x + t)>.|
But that seems impossible because |letgen exp| has to generate
the expression\\|.<let t = exp in body>.| but |letgen| does not have
the |body| yet. The body needs that temporary identifier |.<t>.| that
is supposed to be the result of |letgen| itself. 
Certainly |letgen| cannot generate only a part of a let-expression,
without the |body|.  All generated expressions in
MetaOCaml are well-formed and complete.

The key is to use continuation-passing style (CPS). It was first
pointed out by \cite{Bondorf:92} in the context
of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
generation.

@@

Now, |param_code2 plus one| gives us the desired code\\
|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.

@@

Comparing the code that did let-insertion at the generating stage\\
|let ce = (plus y one) in  plus ce (plus x ce)|\\
with the corresponding code inserting let at the generated code stage\\
|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
clearly shows the difference between the direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
inverted. For another, what used to be an expression that yields
value, |init|, now became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of paranetheses around |fun| increases, the need to add and
then invoke the |k|, continuation argument, become annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

The |do|-notation of Haskell (cire report?) shows that it is possible
to write essentially CPS code in the conventional-looking style. The
|do| notation is the notation for writing monadic code (cite Moggi?).
The benefit of monadic notation is not that it can represent CPS (cite
Filinski), but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad (cite Hudak, Peyton-Jones).

In the present paper, we use the monad that is the CPS monad with
another layer of state. We will discuss the need for the state in
Section XXX. The type of the monad is:
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
\end{code}

The monad is just the function of two arguments: the state (of type
|'s|) and the continuation. The continuation receives the current
state, the value (of type |v|) and yields the answer of the type |w|.
The monad is polymorphic over all these three type parameters. Except
a few primitime monad operations below, all other code treat the monad
as an abstract data type.

Monads represent computations that can have ``effects''. In our case,
we can have ``state effects'' (update of state) and ``control effect''
(which occurs whenever the computation does something else but
invoking its continuation once). As we will see, ``control effects''
in our case manifest themselves as ``code generation effects''.

\begin{code}
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
\end{code}

The two monad operations, |ret| and |bind| are conventional: |ret a|
returns a computation that takes the current state and the
continuation and immediately invokes the continuation passing it the
unmodified state and the value. Obviously, the computation |ret a| has
no control effects. The operation |bind a f| combines a
monad |a| of the type |(s,v,w) monad| and the function |f| of the
type |v -> (s,v1,w) monad| into a bigger |(s,v1,w) monad|. The latter
will execute computation |a|, pass it result to |f| and execute the
latter's computation. The state facet of our monad is handled with two
operations:

\begin{code}
let fetch s k = k s s
let store v s k = k (v::s) ()
\end{code}

|fetch| is the computation that yields the current state and |store v|
is the computation that updates the current state and yields |()|. We
model the state as a list (which is actually a list of polymorphic
variants, as we shall see later).

\begin{code}
let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
\end{code}

The operation |retN a| is the let-insertion operation (whose simpler 
version was called |letgen| before). It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Mention similarity with reset?

Finally, we should be able to run our monad -- that is, perform the
computation of teh monad and get the result -- which in our case is
the code expression. We run the monad by passing it initial state (the
empty list) and the initial continuation |let k0 s v = v|.

We can now re-write our |param_code2| example of the previous section
as

\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  and run m = m [] k0 
  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
\end{code}

That does not seem like the improvement. But here we can introduce the
do notation, which is patterned after the do-namtion of Haskell.
\oleg{Explain our mdo briefly? Refer to the message on the
  Ocaml-list?. Refer to camlp4 and note the implementation}

\begin{code}
let param_code4 plus one =
  let gen x y = mdo { ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) }
  and run m = m [] k0 
  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
\end{code}

The function |param_code4| is totally equivalent to |param_code3| --
the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
preprocessor (with our extension) will transform the |param_code4|
into |param_code3|. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\oleg{Should we rename lif into ifL?}
Now we can build more complex monadic operations.
\begin{code}
let lif test th el = ret .< if .~test then .~th else .~el >.
\end{code}

ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
the connection with the delimited continuations: for future work.

@@

Mention the mdo notation. Mention it into the introduction.
